---
title: "Lab 1 - Nonparametric methods in survival analysis"
author: "Matteo Suardi"
date: "2026-03-29"
output: pdf_document
---

We will work on the dataset described during the lecture: 42 children affected by Acute Lymphoblastic Leukemia
(ALL) in first complete remission (1 CR) randomized (pair matched) to treatment arms (placebo and 6-MP) for remission maintenance (Freirerch et al, 1963). 

Lets load the data frei.txt into R and call the "survival" package.

```{r}
library(survival)
frei <- read.table('Acute lymphoblastic data.txt', na.strings = ".", header = TRUE, row.names = NULL)
names(frei)
head(frei)
```

## Question 1
Working on the pooled sample, estimate and plot the survival function by the Kaplan-Meier method with Greenwood confidence bands.

```{r}
fit <- survfit(Surv(time, status) ~ 1, data = frei)
summary(fit)
names(fit)

par(mar = c(4, 4, 2, 2))
plot(fit, xlab = 'time (weeks)', ylab = 'survival probability')
title('Kaplan Meier estimate (pooled data)')

library(prodlim)
fit.pl <- prodlim(Hist(time, status) ~ 1, data = frei)
plot(fit.pl)
```

## Question 2
For each treatment, estimate and plot the survival function by the KM method.

```{r}
fit <- survfit(Surv(time, status) ~ treat, data = frei)
summary(fit)

plot(fit, col = c(1, 2))
plot(c(0, fit[treat=1]$time), c(1, fit[treat=1]$surv), type = 's', ylim = c(0, 1),
     xlim = c(0, 35), xlab = 'time', ylab = 'survival probability', col = 1)
lines(c(0, fit[treat=2]$time), c(1, fit[treat=2]$surv), type = 's', col = 2)
lines(c(0, fit[treat=1]$time), c(1, fit[treat=1]$lower), type = 's', lty = 2, col = 1)
lines(c(0, fit[treat=1]$time), c(1, fit[treat=1]$upper), type = 's', lty = 2, col = 1)
lines(c(0, fit[treat=2]$time), c(1, fit[treat=2]$lower), type = 's', lty = 2, col = 2)
lines(c(0, fit[treat=2]$time), c(1, fit[treat=2]$upper), type = 's', lty = 2, col = 2)
title('Kaplan Meier estimate according to treatment')

library(prodlim)
fit.pl <- prodlim(Hist(time, status) ~ treat, data = frei)
plot(fit.pl)
```

## Question 3
Contrast in a plot the number of censored subjects and those at risk in the two groups.

```{r}
# At risk
plot(fit[treat=1]$time, fit[treat=1]$n.risk, xlab = 'time',
     ylab = 'number at risk, number of censored',
     type = 's', ylim = c(0, 22), xlim = c(0, 35))
points(fit[treat=1]$time, fit[treat=1]$n.risk)

lines(fit[treat=2]$time, fit[treat=2]$n.risk, type = 's', lwd = 2)
points(fit[treat=2]$time, fit[treat=2]$n.risk, pch = 2)

# Censored subjects (reverse estimation)
fit.rev <- survfit(Surv(time, status == 0) ~ treat, data = frei)

lines(fit.rev[treat=1]$time, fit.rev[treat=1]$n.event, type = 's')
points(fit.rev[treat=1]$time, fit.rev[treat=1]$n.event)

lines(fit.rev[treat=2]$time, fit.rev[treat=2]$n.event, type = 's', lwd = 2)
points(fit.rev[treat=2]$time, fit.rev[treat=2]$n.event, pch = 2)
```

## Question 4
Working on the pooled sample, estimate and plot the cumulative hazard function:
  a. by the Nelson-Aalen estimate
  b. by $-\log(S(t))$
  
```{r}
fit.pool <- survfit(Surv(time, status) ~ 1, data = frei)

# Nelson-Aalen
na.haz <- cumsum(fit.pool$n.event / fit.pool$n.risk)
cbind(fit.pool$time, na.haz)
plot(fit.pool$time, na.haz, type = 's', xlab = 'time', ylab = 'cumulative hazard')

# -log(S(t))
cumhaz <- -log(fit.pool$surv)
plot(fit.pool$time, cumhaz, type = 's', ylim = c(0, 2), xlab = 'time', ylab = 'cumulative hazard')

plot(fit.pool$time, na.haz, type = 's', xlab = 'time (weeks)',
     ylab = 'cumulative hazard', ylim = c(0, 2))
points(fit.pool$time, na.haz)
lines(fit.pool$time, cumhaz, type = 's', col = 2)
points(fit.pool$time, cumhaz, pch = 20, col = 2)
legend("bottomright",
       legend = c("Nelson-Aalen estimate", "-log(S(t))"),
       col = c(1, 2), lty = 1, pch = c(1, 20))
```

## Question 5
Estimate and plot the cumulative hazard in the two treatment groups.

```{r}
na.haz1 <- cumsum(fit[treat=1]$n.event / fit[treat=1]$n.risk)
na.haz2 <- cumsum(fit[treat=2]$n.event / fit[treat=2]$n.risk)
plot(fit[treat=1]$time, na.haz1, xlab = 'time', ylab = 'cumulative hazard',
     type = 's', xlim = c(0, 35))
lines(fit[treat=2]$time, na.haz2, type = 's', lwd = 2)
```

## Question 6
Provide a smoothed estimate of the hazard function of the two treatment groups.

```{r}
haz1 <- fit[treat=1]$n.event / fit[treat=1]$n.risk
haz2 <- fit[treat=2]$n.event / fit[treat=2]$n.risk

plot(fit[treat=1]$time, haz1, xlab = 'time', ylab = 'hazard',
     xlim = c(0, 35), ylim = c(0, 1))
lines(lowess(cbind(fit[treat=1]$time, haz1)), col = 2)
points(fit[treat=2]$time, haz2, lwd = 2)
lines(lowess(cbind(fit[treat=2]$time, haz2)), col = 2)
```

## Question 7
Estimate the failure rate under the hypothesis of constant hazard and interpret in terms of events/person weeks.

```{r}
# Placebo (treat = 1)
tot.time.at.risk <- sum(frei$time[frei$treat == 1])
tot.events       <- sum(frei$status[frei$treat == 1])
f.rate           <- tot.events / tot.time.at.risk
cbind(tot.events, tot.time.at.risk, f.rate)

# 6-MP (treat = 2)
tot.time.at.risk <- sum(frei$time[frei$treat == 2])
tot.events       <- sum(frei$status[frei$treat == 2])
f.rate           <- tot.events / tot.time.at.risk
cbind(tot.events, tot.time.at.risk, f.rate)
```

## Question 8
Test the null hypothesis of absence of treatment effect (6-MP vs PL) using a nonparametric hypothesis test (log-rank test).

```{r}
survdiff(Surv(time, status) ~ treat, data = frei)
```

